Nearest Neighbour Methods for Length of Stay Prediction

### 2015-02-14 papers/paper.tex
Rearrange paragraphs in results/discussion section

Recall that in order for our work to be comparable to that of previous work on
trauma LOS prediction by Dinh et al. \cite{Dinh2013a}, we use the same data set
that they used, and also tested the features they used as one of the manual
feature selection methods.
From Table \ref{tab:overall-tr-acc} we can see that
the best accuracy achieved using their features is 75.06\% with logistic
regression.
It is worth noting that our best accuracy, using $k$-NN with 1
nearest neighbour, shows a statistically significant improvement of 2.75\% upon
the accuracy achieved by Dinh et al.
while using only 11 features (compared to the 19 they used). These 11 features
make up only 14.1\% of all features in the data set, and performed better than
the 11 features selected manually by the domain expert.

@@

Although $k$-NN with 1 nearest neighbour had the highest accuracy (77.81\%) out
of all combinations of classifiers and feature selection methods for this data
set, we should mention that logistic regression came a very close second with
77.7\%, outperforming more sophisticated approaches such as SVM and MLP. It
also achieved a higher AUC, and has the added advantages of being faster to
train, mathematically well-understood and widely applied in medicine.

\subsection{Comparison of Feature Selection Methods}
The best accuracy and AUC results for
each classifier resulted from a reduced feature set obtained through a
particular feature selection method. There was at least one feature selector
for each classifier that managed to outperform the same classifier with
no feature selection. This is likely due to the large number of features in
the original data: after cleaning and preliminary removal of redundant
features, we were still left with over 50. Many of these would still likely
be irrelevant to predicting the class.
Out of 20 combinations of classifiers
and discretisation (the 10 classifiers with and 10 without
discretisation), the C4.5 wrapper method resulted in the best
accuracy for 7 of these combinations, followed by 5 using information gain
and 4 using the correlation coefficient and very few for the other methods.
\texttt{TODO: recount the numbers in this paragraph.}

Focussing our attention on the AUC reveals slightly different patterns. Here we
find that the correlation coefficient and the information gain are the feature
selectors that give the best AUC for most of the classifiers, and perform
equally well when used in conjunction with logistic regression and SVMs.
Interestingly, although the C4.5
wrapper method selects features which allow classifiers to achieve good
accuracy, only in 2 out of the 20 combinations of classifiers did it give the
best AUC result out of all the feature selectors.

Note that regardless of whether we consider accuracy or AUC to be more
important in evaluating classifiers, neither the baseline
feature set from Dinh et al. \cite{Dinh2013a} nor the features suggested by
the domain expert yielded better
results than automatic feature selection in all but \texttt{two?} cases. Neither
feature selection method produced the best AUC result for any classifier.
This is a surprising result because intuitively -- and in the literature -- it
has been stated that in a highly specialised field such as medicine, the domain
experts would be best-equipped to make judgements about which features to
use \cite{Witten2005}, and these should give reasonable results. However,
our results show that in attempting to classify patients according to their
LOS, we should consider both manual and automatic approaches.

Recall that $k$-NN with 1 nearest neighbour and C4.5 wrapper feature
selection produced the best accuracy (77.81\%) for the data set, and
logistic regression with correlation coefficient feature selection at a
threshold of 0.1 gave the best AUC (0.846). These two feature selection
methods selected 11 and 29 features respectively, and the features that
were chosen in common between them were: \texttt{operation}, \texttt{bp},
\texttt{iss}, \texttt{lowerlimbnopelvis}, \texttt{age}, \texttt{lowerlimbany}
and \texttt{icu}. This adds evidence to the work of Dinh et al. who found
that \texttt{iss}, \texttt{age} and \texttt{operation} were important
distinguishers of LOS $\leq$ or $>$ 2 days \cite{Dinh2013a}. We also
confirm the findings of Gabbe et al. who found that \texttt{age},
\texttt{iss} and \texttt{bp} were useful in predicting the length of
hospital stay of blunt trauma patients \cite{Gabbe2005}. Our feature
selection results also
support the findings of McGonigal et al. \cite{McGonigal1993} who used
\texttt{age} and
\texttt{iss} to predict the probability of survival for trauma patients, a
related problem in the same medical domain.
Our observations suggest the use of feature selection not only as a
pre-processing step to improve the accuracy or AUC of a classifier, but also
as a tool used separately that could help extract insights from medical data.

Although we were able to achieve higher accuracy and AUC than the baseline,
our best approaches also have a number of limitations which need to be pointed
out. The first point to note is that using a wrapper method of feature
selection is slower than using a filter approach (or manually picking the
features before training), as the learning algorithm used as the wrapper
needs to be invoked at each of the folds of cross-validation. Additionally,
the $k$-NN algorithm becomes slower as the data set increases in the number
of examples. We did not face such a problem during our work, but this will
need to be addressed for larger data sets. It is also common to use more
than one nearest neighbour, so care needs to be taken to find a suitable number
of neighbours for a given problem. Finally, we achieved an improvement
in AUC of 0.034 but with 29 features, which is still a fairly large
number. The importance of using a fewer number of features is twofold: not
only does this speed up the training time, it also means that we are able to
make predictions without requiring a lot of information about a patient,
which is a crucial consideration if this is to be implemented for use by
physicians.

\subsection{Comparison of Classifiers}
Figures \ref{fig:tr-nothreshold} and \ref{fig:tr-threshold} compare the
accuracy and AUC of all classifiers under each feature selection method.
The methods that do not select features based upon a specified threshold
(manual and baseline as well as C4.5 wrapper and no feature selection)
are graphed in Figure \ref{fig:tr-nothreshold}, and the methods that select
based upon thresholds (correlation and information gain) are graphed
in Figure \ref{fig:tr-threshold}, showing varying accuracy and AUC as the
threshold is changed.

In terms of classification accuracy, we did not find a lot of variation
between all classifiers, with Zero-R producing 66.91\% accuracy and 77.81\%
being the best we achieved -- an improvement of just over 10\%. The AUC of
the 11 classifiers we tested ranged from 0.498 with Zero-R to 0.846 with
logistic regression. Logistic regression, the \textit{de facto} method
used in LOS classification, performed consistently better than the others
across all feature
selection methods, but only when the feature set was not heavily reduced by
the feature selector: in these cases, such as the results from Figure
\ref{fig:tr-nothreshold-cfs-acc} showing classifier performance with features
selected by C4.5 decision trees and nearest
neighbour methods have higher accuracy than logistic regression. This is also
apparent from the graphs in Figure \ref{fig:tr-threshold}: for both
feature selection methods, as we increase the threshold, logistic regression
drops in performance after the initial few thresholds and is outperformed by
C4.5 and the various nearest neighbour classifiers ($k$-NN with 1 and 20
neighbours and K*), as well as our Ranked Distance algorithm.
This is true for both accuracy and AUC.

@@

Although the MLP has been investigated in studies attempting to identify
factors associated with longer LOS or mortality for trauma patients
\cite{Hunter2000,McGonigal1993}, leading to improved results with the neural
network, we did not see the MLP outperform all other classifiers with respect
to accuracy or AUC for any feature selection method for this data set. In fact,
the MLP often performed no better than any of the nearest neighbour
classifiers, which use a straightfoward mechanism of classifying unseen
examples. Even if the MLP had managed to achieve superior accuracy or AUC, the
``black box'' nature of its prediction decisions are a barrier for their
widespread use in medical decision-making.

Nearest neighbour approaches appear to have performed well on this data set in
both accuracy and AUC, especially in comparison to more sophisticated
classifiers and when the feature set was significantly reduced. From the graphs
in Figure \ref{fig:tr-threshold}, we can see that all the nearest neighbours
approaches perform similarly when the number of features is decreased, but our
Ranked Distance algorithm showed a statistically significant improvement over
other NN algorithms. This
improvement is not apparent at smaller feature sets, where using the standard
$k$-NN algorithm with 1 or 20 neighbours performs slightly better.
The performance of NN we observed in our results is
an interesting finding because nearest neighbours have not been commonly
applied in predicting the LOS, but there is intuitive appeal in the
case-by-case nature of how these algorithms classify new examples that should
be explored further. Unlike MLPs and SVMs, prediction decision of nearest
neighbours are more transparent and understandable. However, the major drawback
of NN algorithms is their storage requirement and the computational time
required to search for the nearest neighbours to make the classification.

There is a limitation to our findings that needs to be noted: we did not
attempt to tune the parameters of the classifiers in order to find the ones
that resulted in the best performance, so our statements are based on the
results of the default parameters of each classifier.
MLPs in particular are sensitive to
the network architecture, so it is likely that its accuracy and AUC could be
improved if the parameters were selected to optimise its performance on this
data set.

>>>>

Recall that in order for our work to be comparable to that of previous work on
trauma LOS prediction by Dinh et al. \cite{Dinh2013a}, we use the same data
set and also test the features they used as the baseline feature set.
From Table \ref{tab:overall-tr-acc} we can see that
the best accuracy achieved using their features is 75.06\% with logistic
regression.
It is worth noting that our best accuracy, using $k$-NN with 1
nearest neighbour, shows a statistically significant improvement of 2.75\% upon
the accuracy achieved by Dinh et al.
while using only 11 features (compared to the 19 they used). These 11 features
make up only 14.1\% of all features in the data set, and performed better than
the 11 features selected manually by the domain expert.

@@

Although $k$-NN with 1 nearest neighbour had the highest accuracy (77.81\%) out
of all combinations of classifiers and feature selection methods for this data
set, we should mention that logistic regression came a very close second with
77.7\%, outperforming more sophisticated approaches such as SVM and MLP. It
also achieved a higher AUC, and has the added advantages of being faster to
train and widely applied in medicine.

\subsection{Comparison of Feature Selection Methods}
The best accuracy and AUC results for
each classifier resulted from a reduced feature set obtained through a
particular feature selection method. Out of 20 combinations of classifiers
and discretisation (the 10 classifiers with and 10 without
discretisation), the C4.5 wrapper method resulted in the best
accuracy for 7 of these combinations, followed by 5 using information gain
and 4 using the correlation coefficient and very few for the other methods.
\texttt{TODO: recount the numbers in this paragraph.}

In terms of AUC, we
find that the correlation coefficient and the information gain are the feature
selectors that give the best AUC for most of the classifiers, and perform
equally well when used in conjunction with logistic regression and SVMs.
Interestingly, although the C4.5
wrapper method selects features which allow classifiers to achieve good
accuracy, only in 2 out of the 20 combinations of classifiers did it give the
best AUC result out of all the feature selectors.

Note that regardless of whether we consider accuracy or AUC to be more
important in evaluating classifiers, neither the baseline
feature set from Dinh et al. \cite{Dinh2013a} nor the features suggested by
the expert yielded better
results than automatic feature selection in all but \texttt{two?} cases. Neither
feature selection method produced the best AUC result for any classifier.
This is a surprising result because intuitively -- and in the literature -- it
has been stated that in a highly specialised field such as medicine, the domain
experts would be best-equipped to make judgements about which features to
use \cite{Witten2005}, and these should give reasonable results. However,
our results show that in attempting to classify patients according to their
LOS, we should consider both manual and automatic approaches.

Recall that $k$-NN with 1 nearest neighbour and C4.5 wrapper feature
selection produced the best accuracy (77.81\%) for the data set, and
logistic regression with correlation coefficient feature selection at a
threshold of 0.1 gave the best AUC (0.846). These two feature selection
methods selected 11 and 29 features respectively, and the features that
were chosen in common between them were: \texttt{operation}, \texttt{bp},
\texttt{iss}, \texttt{lowerlimbnopelvis}, \texttt{age}, \texttt{lowerlimbany}
and \texttt{icu}. This adds evidence to the work of Dinh et al. who found
that \texttt{iss}, \texttt{age} and \texttt{operation} were important
distinguishers of LOS $\leq$ or $>$ 2 days \cite{Dinh2013a}. We also
confirm the findings of Gabbe et al. who found that \texttt{age},
\texttt{iss} and \texttt{bp} were useful in predicting the length of
hospital stay of blunt trauma patients \cite{Gabbe2005}. Our feature
selection results also
support the findings of McGonigal et al. \cite{McGonigal1993} who used
\texttt{age} and
\texttt{iss} to predict the probability of survival for trauma patients, a
related problem in the same medical domain.
Our observations suggest the use of feature selection not only as a
pre-processing step to improve the accuracy or AUC of a classifier, but also
as a tool used independently to extract insights from medical data.

\subsection{Comparison of Classifiers}
In terms of classification accuracy, we did not find a lot of variation
between all classifiers, with Zero-R producing 66.91\% accuracy and 77.81\%
being the best we achieved -- an improvement of just over 10\%. The AUC of
the 11 classifiers we tested ranged from 0.498 with Zero-R to 0.846 with
logistic regression. Logistic regression, the \textit{de facto} method
used in LOS classification, performed consistently better than the others
across all feature
selection methods, but only when the feature set was not heavily reduced by
the feature selector: in these cases, such as the results from Figure
\ref{fig:tr-nothreshold-cfs-acc} showing classifier performance with features
selected by C4.5 and nearest
neighbour methods have higher accuracy than logistic regression. This is also
apparent from the graphs in Figure \ref{fig:tr-threshold}: for both
feature selection methods, as we increase the threshold, logistic regression
drops in performance after the initial few thresholds and is outperformed by
C4.5 and the various nearest neighbour classifiers ($k$-NN with 1 and 20
neighbours and K*), as well as our Ranked Distance algorithm.
This is true for both accuracy and AUC.

@@

Although the MLP has been investigated in studies attempting to identify
factors associated with longer LOS or mortality for trauma patients
\cite{Hunter2000,McGonigal1993}, leading to improved results with the neural
network, we did not see the MLP outperform all other classifiers with respect
to accuracy or AUC for any feature selection method for this data set. In fact,
the MLP often performed no better than any of the nearest neighbour
classifiers, which use a straightfoward mechanism of classifying unseen
examples. Even if the MLP had managed to achieve superior accuracy or AUC, the
``black box'' nature of its prediction decisions is a barrier for their
widespread adoption in medical decision-making.

Nearest neighbour approaches appear to have performed well on this data set in
both accuracy and AUC, especially in comparison to more sophisticated
classifiers and when the feature set was significantly reduced. From the graphs
in Figure \ref{fig:tr-threshold}, we can see that all the nearest neighbours
approaches perform similarly when the number of features is decreased, but our
Ranked Distance algorithm showed a statistically significant improvement over
other NN algorithms. This
improvement is not apparent at smaller feature sets, where using the standard
$k$-NN algorithm with 1 or 20 neighbours performs slightly better.
The performance of NN we observed in our results is
an interesting finding because nearest neighbours have not been commonly
applied in predicting the LOS, but there is intuitive appeal in the
case-by-case nature of how these algorithms classify new examples that should
be explored further. Unlike MLPs and SVMs, prediction decision of nearest
neighbours are more transparent and understandable.

Although we were able to achieve higher accuracy and AUC than the baseline,
our best approaches also have a number of limitations which need to be pointed
out. The first point to note is that using a wrapper method of feature
selection is slower than using a filter approach (or manually picking the
features before training), as the learning algorithm used as the wrapper
needs to be invoked at each of the folds of cross-validation. Additionally,
the $k$-NN algorithm becomes slower as the data set increases in the number
of examples. We did not face such a problem during our work, but this will
need to be addressed for larger data sets. It is also common to use more
than one nearest neighbour, so care needs to be taken to find a suitable number
of neighbours for a given problem. Finally, we did not
attempt to tune the parameters of the classifiers in order to find the ones
that resulted in the best performance, so our statements are based on the
results of the default parameters of each classifier.
MLPs in particular are sensitive to
the network architecture, so it is likely that its accuracy and AUC could be
improved if the parameters were selected to optimise its performance on this
data set.
